### 2023-05-03 paper.Rmd
include ARI in the results

The simulation studies were implemented using the statistical programming language R [Version 4.2, @R-base]. The R package `anticlust` [Version 0.6.1, @R-anticlust; @papenberg2020] was used to optimize the anticlustering objectives. The R package `faux` [Version 1.1.0, @R-faux] was used to generate the multivariate normal data sets that were processed during Simulation 2. The R packages `dplyr` [Version 1.0.7, @R-dplyr], `ggplot2` [Version 3.3.5, @R-ggplot2], `papaja` [Version 0.1.1.9001, @R-papaja], `tidyr` [Version 1.1.3, @R-tidyr] and `DescTools` [Version 0.99.45, @R-desctools] were used for data analysis and result presentation. The simulation study is fully reproducible via code and data that has been made accessible in an accompanying OSF repository [@osf2023].

@@

@@

apa_table(
  tab,
  caption = "Results of Simulation study 2",
  note = "The global results of the simulation study aggregated across all conditions. Cells contain information about the average global between-group discrepancy with regard to means, standard deviations, skewness and kurtosis. Lower values indicate less discrepancy, i.e., higher between-group similarity. Each cell value is the result of averaging across 30,000 data points (10,000 data sets $\\times$ $K = 2$, $K = 3$ or $K = 4$).",
  escape = FALSE,
  align = c("r")
)

@@

@@

Table 3 shows the results of the anticlustering application by listing the descriptive statistics (means, standard deviations and skewness) for each of the 9 groups and for each of the 3 features. Table 3 also contains the results after applying k-means anticlustering and diversity anticlustering. It is shown that---up to two decimals---k-plus anticlustering perfectly matched all groups with regard to the features' means and standard deviations, and the skewness was also very evenly matched. K-means anticlustering also perfectly matched the mean values, but showed decreased performance with regard to similarity in standard deviations and skewness. Diversity anticlustering was well-suited to match means, standard deviations and skew, but was slightly outperformed by k-plus. The latter result is not surprising because---unlike k-plus anticlustering---diversity anticlustering does not directly maximize similarity with regard to these criteria, but instead maximizes within-group heterogeneity.

@@

Table 4 illustrates another advantage of k-plus anticlustering over diversity anticlustering: When creating groups of unequal size, diversity anticlustering tends to increase the spread of the data in the largest group in comparison to the other groups. K-plus anticlustering strives for between-group similarity regardless of group size. Unequal group sizes can be requested by passing the different group sizes to the argument `K`. The following code was used to equalize means and variances between four groups using the k-plus approach:

@@

Table 5 illustrates the results that are obtained when using standard[^standardkplus] k-plus anticlustering to divide the stimulus set into 6 groups of size `r nrow(schaper2019) / 6` each. Even though the means and standard deviations are no longer perfectly matched between groups, arguably they are similar enough for any practical purposes.

@@

Anticlustering is usually implemented by reversing an objective function used in classical cluster analysis. Whereas cluster analysis seeks homogeneity within groups and separation between groups, the reversal results in heterogeneity within groups and similarity between groups. While tightly linked in a mathematical sense, an important semantic difference is that cluster analysis is concerned with finding "true" structure in the data, e.g., by defining a taxonomy on naturally occurring groups [@everitt1979]. Anticlustering instead imposes structure on the data, while the anticlusters themselves do not represent any underlying "true" pattern. Correspondingly, different---but similarly adequate---anticlustering partitions tend to look very different, while good clustering solutions tend to look similar. A good clustering solution assigns similar elements into the same cluster and dissimilar elements into different clusters. A good anticlustering solution does not depend on assigning individual items either to the same or a different group, independent of their pairwise dissimilarity.[^dispersion] This pattern was observed in SImulation 2, where it was shown that partitions obtaining high between-group similarity according to different criteria were not at all similar; the ARI was usually close to 0. For different clustering methods, there is usually a considerable amount of overlap between partitions. This  distinction between cluster analysis and anticlustering can also be clarified using Figure 1, where we observe a high concentration of suitable anticlustering partitions but only few suitable clustering solutions. An astonishing `r mean(df$kmeans / max(df$kmeans) >.999) * 100`% of all partitions are within 0.1% of the global maximum in $\mathit{SSE}$, i.e., the best k-means anticlustering solution. Among these `r sum(df$kmeans / max(df$kmeans) >.999)` partitions, every possible pair of elements is sometimes assigned to the same group and sometimes not. In contrast, there are only `r sum(min(df$kmeans)/ df$kmeans > .90)` partitions (`r mean(min(df$kmeans)/ df$kmeans > .90)*100`%) that approximate the global minimum in $\mathit{SSE}$ (i.e., the best k-means clustering solution) by even 10%. Thus, even though the problems are based on the same objective function, cluster analysis and anticlustering exhibit a stark discrepancy with regard to the solution space of partitions. Interestingly, the k-plus objective introduces an additional distinction between cluster analysis and anticlustering. Reversing the k-plus anticlustering objective no longer yields a useful clustering interpretation, because identifying groups that are different with regard to higher order moments does not constitute a clustering. Thus, even though k-plus extends the classical k-means clustering criterion, it is a pure anticlustering objective.

>>>>

The simulation studies were implemented using the statistical programming language R [Version 4.2, @R-base]. The R package `anticlust` [Version 0.6.1, @R-anticlust; @papenberg2020] was used to optimize the anticlustering objectives. The R package `faux` [Version 1.1.0, @R-faux] was used to generate the multivariate normal data sets that were processed during Simulation 2. The R packages `dplyr` [Version 1.0.7, @R-dplyr], `ggplot2` [Version 3.3.5, @R-ggplot2], `papaja` [Version 0.1.1.9001, @R-papaja], `tidyr` [Version 1.1.3, @R-tidyr], `fossil` [Version 0.4.0, @R-fossil], and `DescTools` [Version 0.99.45, @R-desctools] were used for data analysis and result presentation. The simulation study is fully reproducible via code and data that has been made accessible in an accompanying OSF repository [@osf2023].

@@

As a secondary purpose of Simulation 2, I investigated the degree to which partitions identified by the different methods showed similarity. Specifically, I investigated whether different anticlustering solutions showed any overlap regarding the placement of individual elements in groups. In cluster analysis, high quality partitions according to different objectives tend to overlap to a large degree, but it is an open question whether this result also holds for different anticlustering partitions. To tackle this issue, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each data set and for each $K$, i.e., 30,000 times.

@@

apa_table(
  tab,
  caption = "Results of Simulation 2.",
  note = "The global results of the simulation study aggregated across all conditions. Cells contain information about the average global between-group discrepancy with regard to means, standard deviations, skewness and kurtosis. Lower values indicate less discrepancy, i.e., higher between-group similarity. Each cell value is the result of averaging across 30,000 data points (10,000 data sets $\\times$ $K = 2$, $K = 3$ or $K = 4$).",
  escape = FALSE,
  align = c("r")
)

@@

ARIs <- as.matrix(read.table("Simulation_2/global_results_ari.csv", sep = ","))
colnames(ARIs) <- c(
  "diversity", "k-means", "k-plus", "k-plus generalized", "random"
)
rownames(ARIs) <- colnames(ARIs)
ARIs <- round(ARIs, 3)
ARIs[upper.tri(ARIs)] <- ""

# removing leading 0
ARIs <- gsub("0.", ".", ARIs, fixed = TRUE)

apa_table(
  ARIs,
  caption = "Average pairwise ARI returned by the competing methods in Simulation 2.",
  note = "Each cell value is the result of averaging across 30,000 data points (10,000 data sets $\\times$ $K = 2$, $K = 3$ or $K = 4$).",
  escape = FALSE,
  align = c("l")
)

Table 3 illustrates the degree of overlap between anticlustering partitions that were identified in Simulation 2. Overall, it is shown that there was hardly any similarity between anticlustering partitions according to the different criteria. The average ARI was almost 0 in all cases---indicating no similarity of partitions whatsoever. This constitutes an interesting and novel result, illustrating that high between-group similarity can be achieved via very different partitionings.

@@

Table 4 shows the results of the anticlustering application by listing the descriptive statistics (means, standard deviations and skewness) for each of the 9 groups and for each of the 3 features. Table 4 also contains the results after applying k-means anticlustering and diversity anticlustering. It is shown that---up to two decimals---k-plus anticlustering perfectly matched all groups with regard to the features' means and standard deviations, and the skewness was also very evenly matched. K-means anticlustering also perfectly matched the mean values, but showed decreased performance with regard to similarity in standard deviations and skewness. Diversity anticlustering was well-suited to match means, standard deviations and skew, but was slightly outperformed by k-plus. The latter result is not surprising because---unlike k-plus anticlustering---diversity anticlustering does not directly maximize similarity with regard to these criteria, but instead maximizes within-group heterogeneity.

@@

Table 5 illustrates another advantage of k-plus anticlustering over diversity anticlustering: When creating groups of unequal size, diversity anticlustering tends to increase the spread of the data in the largest group in comparison to the other groups. K-plus anticlustering strives for between-group similarity regardless of group size. Unequal group sizes can be requested by passing the different group sizes to the argument `K`. The following code was used to equalize means and variances between four groups using the k-plus approach:

@@

Table 6 illustrates the results that are obtained when using standard[^standardkplus] k-plus anticlustering to divide the stimulus set into 6 groups of size `r nrow(schaper2019) / 6` each. Even though the means and standard deviations are no longer perfectly matched between groups, arguably they are similar enough for any practical purposes.

@@

Anticlustering is usually implemented by reversing an objective function used in classical cluster analysis. Whereas cluster analysis seeks homogeneity within groups and separation between groups, the reversal results in heterogeneity within groups and similarity between groups. While tightly linked in a mathematical sense, an important semantic difference is that cluster analysis is concerned with finding "true" structure in the data, e.g., by defining a taxonomy on naturally occurring groups [@everitt1979]. Anticlustering instead imposes structure on the data, while the anticlusters themselves do not represent any underlying "true" pattern. Correspondingly, different---but similarly adequate---anticlustering partitions tend to look very different, as shown in Simulation 2. In contrast, good clustering solutions tend to look similar because a good clustering is characterized by assigning similar elements into the same cluster and dissimilar elements into different clusters. A good anticlustering solution however does not depend on assigning individual items either to the same or a different group, independent of their pairwise dissimilarity.[^dispersion] This distinction can also be clarified using Figure 1, where we observe a high concentration of suitable anticlustering partitions but only few suitable clustering solutions. An astonishing `r mean(df$kmeans / max(df$kmeans) >.999) * 100`% of all partitions are within 0.1% of the global maximum in $\mathit{SSE}$, i.e., the best k-means anticlustering solution. In contrast, there are only `r sum(min(df$kmeans)/ df$kmeans > .90)` partitions (`r mean(min(df$kmeans)/ df$kmeans > .90)*100`%) that approximate the global minimum in $\mathit{SSE}$ (i.e., the best k-means clustering solution) by even 10%. Thus, even though the problems are based on the same objective function, cluster analysis and anticlustering exhibit a stark discrepancy with regard to the solution space of partitions. Interestingly, the k-plus objective introduces an additional distinction between cluster analysis and anticlustering. Reversing the k-plus anticlustering objective no longer yields a useful clustering interpretation, because identifying groups that are different with regard to higher order moments does not constitute a clustering. Thus, even though k-plus extends the classical k-means clustering criterion, it is a pure anticlustering objective.
